Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-01-17 ParisEM.tex
Modifications in proofs and simulations

At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are initialized by setting $\xi^{\ell}_0=x_0$. Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = g_0 (\xi^{\ell}_0)$. 

@@

In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown. Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable. Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011},
an extension of the PaRIS algorithm may be developed by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
\[
\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta)\middle| \mathcal{F}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
\]
where 
\[
\mathcal{F}_{k+1}^N = \sigma\left\{Y_{0:k+1};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k+1\right\}\eqsp.
\]
In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR:unknownq}. Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
\begin{lemma}
\label{lem:AR:unbiased}
Let $0\le k\le n-1$.  For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k}^N(i,\cdot)$. 
\end{lemma}
\begin{proof}
See Appendix \ref{sec:append:proofs}.
\end{proof}
\begin{algorithm}[H]
\caption{Random weight accept-reject backward sampling}
\begin{algorithmic}
\FORALL{$i \in 1,\dots, N$}
\FORALL{$\ell \in 1,\dots, \widetilde N$} 
\STATE \textbf{\sc Sampling Step} \\
Sample independently $
\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\omega_{k}^1,\dots,\omega_{k}^N\}$.
\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
\STATE Set $J_k^{i,\ell} = J$.
\ELSE 
\STATE Return to \textbf{\sc Sampling Step}
\ENDIF
\ENDFOR
\ENDFOR
\end{algorithmic}
\label{alg:AR:unknownq}
\end{algorithm}

@@

\begin{hypA}
\label{assum:boundalgo}
$\underset{k\geq 1}{\sup}|\vartheta_k|_{\infty} < \infty$, $\underset{k\geq 1}{\sup}|p_k|_{\infty} < \infty$ and $\underset{k\geq 1}{\sup}|\widehat{\omega}_{k}|_{\infty} < \infty$, where
\[
\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{\qk}(x,x';z)g_{k+1}(x')}{\vartheta_{k+1}(x) p_{k} (x,x')}\eqsp.
\]
\end{hypA}
%The following Lemma is proved in \cite[Lemma~11]{olsson:westerborn:2016}.
%\begin{lemma}
%\label{lem:smooth:rec}
%Forward smoothing
%\end{lemma}

\begin{lemma}
\label{lem:iid}
For all $0\le k \le n-1$, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and%For all bounded measurable function $h$ on $\mathbb{R}^d$,
\[
\mathbb{E}\left[\tilde{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k+1}(\cdot,x)\right\}\rmd x\right]\eqsp.
\]
\end{lemma}

@@

\begin{proposition}
Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold and that for all $1\le k\le n$, $\mathrm{osc}(h_k)<+\infty$. For all $0\le k\le n$, there exist $b_k,c_k>0$ such that for all $N\ge 1$ and all $\varepsilon\in\mathbb{R}_+^\star$,
\[
\mathbb{P}\left(\left|\phi_k^N[\tau_k] - \phi_k\left[T_kh_k\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
\end{proposition}

\begin{proof}
Write
\[
\phi_{k+1}^N[\tau_{k+1}] - \phi_{k+1}\left[T_{k+1}h_{k+1}\right] = \frac{a_N}{b_N}\eqsp,
\]
where $a_N = N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i$. By Lemma~\ref{lem:iid}, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
\[
\left|\widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\right| \le 2|\widehat{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
\]
Therefore, by Hoeffing inequality,
\[
\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
On the other hand,
\[
\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,%\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}
\]
where
\[
\Upsilon_k(x_k) = \int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left(\tau_k(x_k) + h_{k+1}(x_k,x) - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\rmd x\eqsp.
\]
By \cite[Lemma~11]{olsson:westerborn:2016}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by Proposition~\ref{prop:filter} that 
\[
\mathbb{P}\left(\left|\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
Then,
\[
\mathbb{P}\left(\left|a_N\right|\ge \varepsilon\right) \le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
Similarly, as $b_N \le |\widehat{\omega}_k|_{\infty}$, by Hoeffding inequality,
\[
\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
Note that
\[
\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\rmd x\right]\eqsp.%\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}
\]
By Proposition~\ref{prop:filter},
\[
\mathbb{P}\left(\left|\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]-\phi_k\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\rmd x\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
The proof is completed using Lemma~\ref{lem:hoeffding:ratio}.
\end{proof}

\begin{proposition}
\label{prop:filter}
Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold. Then, for all $0\le k\le n$,
there exist $b_k, c_k>0$ such that for all $N\ge 1$, all $\varepsilon > 0$, and all measurable function $h$ on $\mathsf{X}$,
\begin{equation*}
\mathbb{P}\left( \left|\phi_k^N[h] - \phi_k\left[h\right]\right| \geq \varepsilon \right) \leq b_k \mathrm{exp}\left(-c_k N \epsilon^2\right)\eqsp.
\end{equation*}
\end{proposition}

\begin{proof}
\textcolor{red}{On l'ecrit peut-etre... }
The proof is a straightforward extension of \cite{douc:garivier:moulines:olsson:2011}. 
\end{proof}

@@

\section{Numerical experiments}
\label{sec:exp}
This section investigates the performance of the proposed algorithm. The different tuning parameters are:
\begin{enumerate}[i)]
\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$.
The algorithm is compared to the fixed-lag particle smoother of \cite{olsson:strojby:2011}. To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated. Instead of the Ozaki discretization, the following bridges process are considered:
\begin{enumerate}[i)]
\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
\item a guided bridge process, see \cite{}.
\paragraph{Choice of $\mu$}
A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
% An important feature of this estimator is that under the assumptions of this section it may be bounded uniformly in $x,y$, $\sup_{x,y\in\mathbb{R}^d}  q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) \le M_\theta$ where:
% M_{\theta} = \sup_{x,y\in\mathbb{R}^d} \left(\widetilde{q}^{t}_{\theta}(x,y)\mathrm{exp}\left\{ H_{\theta}(y) - H_{\theta}(x)\right\}\right)\mathrm{exp}\left\{\ell(\theta)t\right\}\eqsp.
\paragraph{Choice of bridge process}
The approximation of $q_{\theta}^{\Delta t_k}$ described in the paper may be compared to the approximation provided by other bridges with affine drift functions. Following \cite{}, let $(Y^{\beta}_t)_{t_{k-1}\le t\le t_k}$ be governed by the following stochastic differential equation for $t_{k-1}\le t\le t_k$:
Y^{\beta}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\beta}_t = \beta\rmd t + \rmd W_t\eqsp.
As suggestd by \cite{}, a guided process between $t_{k-1}$ and $t_k$ may alo be used:
Y^{\mathsf{guid}}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\mathsf{guid}}_t = \frac{X_k - Y^{\mathsf{guid}}_t}{t_k-t}\rmd t + \rmd W_t\eqsp.
Both processes lead to estimators of $q_{\theta}^{\Delta t_k}$ following the same steps as in Section~\ref{}.

>>>>

At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$. 

@@

In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown. Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable. Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011},
an extension of the PaRIS algorithm may be developed by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
\[
\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta)\middle| \mathcal{F}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
\]
where, for all $0\le k \le n$, 
\[
\mathcal{F}_{k}^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k\right\}\eqsp.
\]
In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the accept reject mechanism given in Algorithm~\ref{alg:AR:unknownq}. Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
\begin{lemma}
\label{lem:AR:unbiased}
Let $0\le k\le n-1$.  For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k}^N(i,\cdot)$. 
\end{lemma}
\begin{proof}
See Appendix \ref{sec:append:proofs}.
\end{proof}
\begin{algorithm}[H]
\caption{Random weight accept-reject backward sampling}
\begin{algorithmic}
\FORALL{$i \in 1,\dots, N$}
\FORALL{$\ell \in 1,\dots, \widetilde N$} 
\STATE \textbf{\sc Sampling Step} \\
Sample independently $
\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\omega_{k}^1,\dots,\omega_{k}^N\}$.
\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
\STATE Set $J_k^{i,\ell} = J$.
\ELSE 
\STATE Return to \textbf{\sc Sampling Step}
\ENDIF
\ENDFOR
\ENDFOR
\end{algorithmic}
\label{alg:AR:unknownq}
\end{algorithm}

@@

\begin{hypA}
\label{assum:boundalgo}
$\underset{k\geq 1}{\sup}|\vartheta_k|_{\infty} < \infty$, $\underset{k\geq 1}{\sup}|p_k|_{\infty} < \infty$ and $\underset{k\geq 1}{\sup}|\widehat{\omega}_{k}|_{\infty} < \infty$, where
\[
\widehat{\omega}_{0}(x) = \frac{\chi(x)g_0(x)}{\eta_0(x)} \quad\mbox{and for}\; k\ge1\quad\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{\qk}(x,x';z)g_{k+1}(x')}{\vartheta_{k+1}(x) p_{k} (x,x')}\eqsp.
\]
\end{hypA}
%The following Lemma is proved in \cite[Lemma~11]{olsson:westerborn:2016}.
%\begin{lemma}
%\label{lem:smooth:rec}
%Forward smoothing
%\end{lemma}

\begin{lemma}
\label{lem:iid}
For all $0\le k \le n-1$, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and%For all bounded measurable function $h$ on $\mathbb{R}^d$,
\[
\mathbb{E}\left[\widehat{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int q_{k}(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k+1}(\cdot,x)\right\}\rmd x\right]\eqsp.
\]
\end{lemma}

@@

\begin{proposition}
Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold and that for all $1\le k\le n$, $\mathrm{osc}(h_k)<+\infty$. For all $0\le k\le n$ and all $\widetilde{N}\ge 1$, there exist $b_k,c_k>0$ such that for all $N\ge 1$ and all $\varepsilon\in\mathbb{R}_+^\star$,
\[
\mathbb{P}\left(\left|\phi_k^N[\tau_k] - \phi_k\left[T_kh_k\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
\end{proposition}

\begin{proof}
The results is proved by induction. At time $k=0$, the result holds using that for all $1\le i \le N$, $\rho_0^i = 0$ and the convention $T_0[h_0] =0$. In addition, $\phi_0^N$ is a standard importance sampler estimator of $\phi_0$ with $\omega_0^i\le |\widehat{\omega}_0|_{\infty}$ so that for any bounded function $h$ on $\mathsf{X}$,
\[
\mathbb{P}\left(\left|\phi_0^N[h] - \phi_0\left[h\right]\right|\ge \varepsilon\right)\le b_0\exp\left(-c_0N\varepsilon^2\right)\eqsp.
\]
Assume the results holds for $k\ge 1$ and write
\[
\phi_{k+1}^N[\tau_{k+1}] - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right] = a_N/b_N\eqsp,
\]
where $a_N = N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i$. By Lemma~\ref{lem:iid}, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
\[
\left|\widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\right| \le 2|\widehat{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
\]
Therefore, by Hoeffding inequality,
\[
\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
On the other hand,
\[
\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,
\]
where
\[
\Upsilon_k(x_k) = \int q_{k}(\cdot,x)g_{k+1}(x)\left(\tau_k(x_k) + h_{k+1}(x_k,x) - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\rmd x\eqsp.
\]
By \cite[Lemma~11]{olsson:westerborn:2016}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by the induction assumption that 
\[
\mathbb{P}\left(\left|\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
Then,
\[
\mathbb{P}\left(\left|a_N\right|\ge \varepsilon\right) \le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
Similarly, as $b_N \le |\widehat{\omega}_k|_{\infty}$, by Hoeffding inequality,
\begin{multline*}
\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) \\
= \mathbb{E}\left[\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\end{multline*}
Note that
\[
\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\eqsp.
\]
By  the induction assumption,
\[
\mathbb{P}\left(\left|\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]-\phi_k\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
The proof is completed using Lemma~\ref{lem:hoeffding:ratio}.
\end{proof}

@@

\section{Numerical experiments}
\label{sec:exp}
%This section investigates the performance of the proposed algorithm. The different tuning parameters are:
%\begin{enumerate}[i)]
%\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
%\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
%\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
%\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
%\end{enumerate}
%The performance of the algorithm is investigated on two known models based on diffusion process.

\subsection*{Sine model} 
The algorithm is compared to the fixed-lag particle smoother of \cite{olsson:strojby:2011}.

%To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated. Instead of the Ozaki discretization, the following bridges process are considered:
%\begin{enumerate}[i)]
%\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
%\item a guided bridge process, see \cite{}.
%\end{enumerate}
%\paragraph{Choice of $\mu$}
%A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
%\begin{multline*}
%q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
%\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
%\end{multline*}

% An important feature of this estimator is that under the assumptions of this section it may be bounded uniformly in $x,y$, $\sup_{x,y\in\mathbb{R}^d}  q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) \le M_\theta$ where:
% \[
% M_{\theta} = \sup_{x,y\in\mathbb{R}^d} \left(\widetilde{q}^{t}_{\theta}(x,y)\mathrm{exp}\left\{ H_{\theta}(y) - H_{\theta}(x)\right\}\right)\mathrm{exp}\left\{\ell(\theta)t\right\}\eqsp.
% \]
%
%\paragraph{Choice of bridge process}
%The approximation of $q_{\theta}^{\Delta t_k}$ described in the paper may be compared to the approximation provided by other bridges with affine drift functions. Following \cite{}, let $(Y^{\beta}_t)_{t_{k-1}\le t\le t_k}$ be governed by the following stochastic differential equation for $t_{k-1}\le t\le t_k$:
%\[
%Y^{\beta}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\beta}_t = \beta\rmd t + \rmd W_t\eqsp.
%\]
%As suggestd by \cite{}, a guided process between $t_{k-1}$ and $t_k$ may alo be used:
%\[
%Y^{\mathsf{guid}}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\mathsf{guid}}_t = \frac{X_k - Y^{\mathsf{guid}}_t}{t_k-t}\rmd t + \rmd W_t\eqsp.
%\]
%Both processes lead to estimators of $q_{\theta}^{\Delta t_k}$ following the same steps as in Section~\ref{}.
\subsection*{Log-growth model}
Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and \cite{olsson:westerborn:2016}, the performance of the proposed algorithm are also illustrated with the log-growth model defined by:
\begin{equation}
\rmd Z_t = \kappa Z_t\left(1-\frac{Z_t}{\gamma}\right)\rmd t + \sigma Z_t \rmd W_t,~~Z_0=z_0. \label{eq:LG:SDE}
\end{equation}
In order to use the exact algorithms of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}, we consider  \eqref{eq:LG:SDE} after the Lamperti transform, i.e., the process defined by $X_t=\eta(Z_t)$, with $\eta(z) := -\log (z)/\sigma$,  which satisfies the following SDE:
\begin{equation}
\rmd X_t = \overbrace{\left( \frac{\sigma}{2} -  \frac{\kappa}{\sigma} + \frac{\kappa}{\gamma\sigma}\exp\left(-\sigma X_t\right)\right)}^{:=\alpha(X_t)}\rmd t +\rmd W_t,~~X_0=x_0=\eta(z_0).\label{eq:Lamp:LG}
\end{equation}
In this case, the conditions of the Exact Algorithm~2 defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are satisfied, as for any $m \in \mathbb{R}$ there exists $\mathsf{U}(m)$ such that for all $x\ge m$, $\psi(x):=\alpha^2(x)+\alpha'(x) \leq \mathsf{U}_m$. Moreover, $\psi$ is lower bounded uniformly by  $\mathsf{L}$.  Then, GPE estimators may be computed by simulating a the minimum of a OU bridge, and simulating OU bridges conditionally to this minimum.  However, to the best of our knowledge, no general techniques exist to do so.  We therefore apply the proposed algorithm, but with a GPE estimator computed with Bessel bridges, as proposed by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.

The process solution to \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{0\le k\le 50}$ defined as:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:LG}\eqsp,
\end{equation}
where the $(\varepsilon_k)_{0\le k \le 50}$ are i.i.d. $ \mathcal{N}(0,\sigma^2_{obs})$.
The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50). We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:LG}.
This estimation is performed 300 times, using $N=400$ particles, with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.  Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite{olsson:strojby:2011} (Appendix B) based on 30 independent Monte Carlo simulations.

The intermediate quantity of the EM algorithm is also estimated with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:LG}.
The results for the fixed lag technique are similar to the ones presented in \cite{olsson:strojby:2011} (Figure 1). For small lags, the variance of the estimates is small, but the estimation is highly biased. The bias rapidly decreases as the lag increases, together with a  great variance increase.  The PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the variance of the fixed lag when the lag is one. 
%It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property. 
%As the PaRIS algorithm is always unbiased, there is no need to consider the forgetting property of a given model in the approximation of the EM intermediate quantity. 
\begin{figure}[p]
\centering
\includegraphics[width=0.8\textwidth]{obs}
\caption{One realisation of the process $(X_t)_{t\geq 0}$, solution of the SDE \eqref{eq:Lamp:LG} is represented in white. The observations of $Y$ at times $t_0=0,\dots,t_{50}=100$, satisfying equation \eqref{eq:obs:LG},  are represented in black.}
\label{fig:obs:LG}
\end{figure}
\begin{figure}[p]
\centering
\includegraphics[width=0.8\textwidth]{res_Estep_LG}
\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the PaRIS algorithm for diffusion process proposed in this work, using 300 replicates. The whiskers represent the extent of the 95\% central values. The dot represents the empirical mean over the 300 replicates. The dotted line shows the reference value, computed using the unbiased PaRIS algorithm with $N=10000$ particles.}
\label{fig:res:LG}
\end{figure}
